Updated Constraints on Self-Interacting Dark Matter from Supernova 1987A

### 2017-05-15 draft.tex
Added references and a few comments.

An overwhelming preponderance of observational evidence indicates that a form of nonrelativistic, nonbaryonic, 
dark matter constitutes the majority of mass in the Universe and drives the formation of cosmic structure. 
The pace of the quest to identify the dark matter is accelerating on many fronts. Weakly-interacting massive 
particles (WIMPs) have received the most attention as dark matter candidates (see Ref.~\cite{jungman_etal96} for a review). 
Dark matter particles that interact with standard model particles only weakly, while interacting among themselves 
much more strongly have been studied as an alternative to WIMP scenarios in numerous contexts 
\cite{carlson_etal92,deLaix_etal95,atrio-barandela_davidson97,spergel_steinhardt00,hogan_dalcanton00,mohapatra_teplitz00,
dave_etal01,hisano_etal04,hisano_etal05,pospelov_etal08,arkani-hamed_etal08a,lattanzi_silk08,ackerman_etal09,feng_etal09,
kong_etal15} \arz{Cameron: Please review the literature and add some more recent references here. 
I know that Annika Peter has a few papers and Jonathan Feng probably has some more. 
Ensuring that you are up to date on these papers is important.} \arz{As far as I can see, no new 
references were added here or in any other relevant spots in the paper. Please do a literature search. At the very least, 
I know of papers by Annika Peter and Jonathan Feng that may be relevant here.}
and constraints on self-interacting dark matter (SIDM) models have been explored by many authors \cite{yoshida_etal00,gnedin_ostriker01,miralda-escude02,randall_etal08,kamionkowski_profumo08,zentner09,robertson_zentner09,pieri_etal09,spolyar_etal09,finkbeiner_etal09,slatyer_etal09,bramante_etal14,albuquerque_etal14,kaplinghat_etal14,chen_etal14,feng_etal16,catena_widmark16}. In this paper, we revisit and update astrophysical constraints on SIDM models from supernova cooling.

SIDM models in which large self-interaction cross sections are mediated by sufficiently light 
bosons ($M \lesssim 100$~GeV) can be constrained astrophysically using supernovae, particularly 
SN1987A. Light gauge bosons will be produced within the hot supernova core, primarily through 
bremsstrahlung, and radiated. This non-standard energy loss mechanism can result in an energy loss rate 
from the supernova core that is inconsistent with observations of neutrinos from SN1987A, 
analogous to the classic constraint on axions \cite{turner88,raffelt96_book}. 
This effect has already been exploited by Dent et al. \cite{dent_etal12}, Rrapaj and Reddy \cite{rrapaj_reddy16}, and 
Hardy and Lasenby \cite{hardy_lasenby17} \arz{Cameron, add others of which you may be aware.}
to constraint dark electromagnetism models in which the new gauge boson, the so-called dark photon, 
is kinetically mixed with the standard model photon. The SN1987A constraint places a limit on the mixing parameter.

@@

There are two processes to consider in order to estimate the rate of energy loss via $\Aprime$ bremsstrahlung. The first is  
proton-proton (pp) scattering with the bremsstrahlung of the dark photon off the proton; $p+p \rightarrow p+p+\Aprime$. 
The second is proton-neutron (pn) scattering with bremsstrahlung off of either the proton or the charged pion; 
$p+n \rightarrow p+n+\Aprime$. We estimate the rates for these processes using the one-pion exchange (OPE) 
approximation for nucleon interactions. In the pp case, there are eight tree-level diagrams, 
with the emission of the $\Aprime$ from each of the external legs. One of these diagrams is shown 
in Fig.~\ref{fig:ppdiagram}; the remaining seven diagrams come from placing the radiated $\Aprime$ 
on each of the other three protons and then, for each of these, interchanging the outgoing momenta. 
For the pn case, there are five diagrams, four of which are analogous to the pp diagram shown in 
Fig.~\ref{fig:ppdiagram}. The fifth diagram, shown in Fig.~\ref{fig:npdiagram}, corresponds to 
emission of the $\Aprime$ from the exchanged, charged pion.

@@

\begin{figure*}[th]
\includegraphics[width=14cm]{compiledbounds_edited.pdf}
\caption{
Constraints on dark photon models. Our primary result is the excluded region 
in the $\varepsilon$-$m_{\mathrm{A'}}$ plane shaded in blue. The red curve is our attempts to reproduce exactly Ref.~\cite{dent_etal12}, while the black curve is Ref.~\cite{dent_etal12} but with the correct luminosity constraint.  The purple curve is using the SRA fixing the cross section using nuclear scattering data, while the green curve uses the soft radiation approximation of the OPE result.  \arz{The labels in the figure legend appear to be incorrect now.}
}
\label{fig:constraint}
\end{figure*}

@@

Finally, we note that the interior temperature and density within the progenitor of SN1987A remain 
uncertain at least at the level of tens of percent \citep[e.g.][]{perego15,farmer16}. This is well worth 
noting because the constraints one derives using the techniques in this work or similar techniques 
are quite sensitive to the properties of the SN1987A core, particularly the core temperature. Though 
we do not show this result explicitly, we have replicated our calculations for shifts in core temperature 
of $\pm 5\, \mathrm{MeV}$ and find that the boundaries of the excluded region shift by a factor of $\sim 2-3$ at low 
mass ($m_{\mathrm{A}'} \lesssim T$ where $T = 30\, \mathrm{MeV}$ in our fiducial calculation, following 
work or previous authors) and considerably more at high mass. 
We conclude that the differences among theoretical techniques for 
computing the excluded region (OPE, EFT, SRA, ...) lead to uncertainties that are of the same order or 
smaller than the uncertainty in the excluded region induced by our limited knowledge of the 
properties of the SN1987A progenitor interior. The excluded regions we and other authors delineate 
should be considered to be uncertain by a factor or $\sim 3$. Refining the theoretical approach to this 
problem is therefore unwarranted until a breakthrough in our understanding of the interior of the 
SN1987A progenitor is achieved.

>>>>

An overwhelming preponderance of observational evidence indicates that a form of nonrelativistic, nonbaryonic, 
dark matter constitutes the majority of mass in the Universe and drives the formation of cosmic structure. 
The pace of the quest to identify the dark matter is accelerating on many fronts. Weakly-interacting massive 
particles (WIMPs) have received the most attention as dark matter candidates (see Ref.~\cite{jungman_etal96} for a review). 
Dark matter particles that interact with standard model particles only weakly, while interacting among themselves 
much more strongly have been studied as an alternative to WIMP scenarios in numerous contexts 
\cite{carlson_etal92,deLaix_etal95,atrio-barandela_davidson97,spergel_steinhardt00,hogan_dalcanton00,mohapatra_teplitz00,
dave_etal01,hisano_etal04,hisano_etal05,pospelov_etal08,arkani-hamed_etal08a,lattanzi_silk08,ackerman_etal09,feng_etal09,
kong_etal15} \arz{Cameron: Please review the literature and add some more recent references here. 
I know that Annika Peter has a few papers and Jonathan Feng probably has some more. 
Ensuring that you are up to date on these papers is important.} \arz{As far as I can see, no new 
references were added here or in any other relevant spots in the paper. Please do a literature search. At the very least, 
I know of papers by Annika Peter and Jonathan Feng that may be relevant here.} \akl{I added these: }\cite{Buckley:2009in,Boddy:2014yra,Boddy:2014qxa,Boddy:2016bbu}
and constraints on self-interacting dark matter (SIDM) models have been explored by many authors \cite{yoshida_etal00,gnedin_ostriker01,miralda-escude02,randall_etal08,kamionkowski_profumo08,zentner09,robertson_zentner09,pieri_etal09,spolyar_etal09,finkbeiner_etal09,slatyer_etal09,bramante_etal14,albuquerque_etal14,kaplinghat_etal14,chen_etal14,feng_etal16,catena_widmark16} \akl{I added these:} \cite{Markevitch:2003at,Zavala:2012us,Rocha:2012jg,Peter:2012jh,Kahlhoefer:2013dca,Elbert:2014bma,Feng:2015hja,DelNobile:2015uua,Feng:2016ijc,Dooley:2016ajo,Kim:2016ujt,Bringmann:2016din}. 
In this paper, we revisit and update astrophysical constraints on SIDM models from supernova cooling.

SIDM models in which large self-interaction cross sections are mediated by sufficiently light 
bosons ($M \lesssim 100$~GeV) can be constrained astrophysically using supernovae, particularly 
SN1987A. Light gauge bosons will be produced within the hot supernova core, primarily through 
bremsstrahlung, and radiated. This non-standard energy loss mechanism can result in an energy loss rate 
from the supernova core that is inconsistent with observations of neutrinos from SN1987A, 
analogous to the classic constraint on axions \cite{turner88,raffelt96_book}. 
This effect has already been exploited by Dent et al. \cite{dent_etal12}, Rrapaj and Reddy \cite{rrapaj_reddy16}, and 
Hardy and Lasenby \cite{hardy_lasenby17} \arz{Cameron, add others of which you may be aware.}
to constraint dark electromagnetism models in which the new gauge boson, the so-called dark photon, 
is kinetically mixed with the standard model photon. \akl{I added these:} See also \cite{Kazanas:2014mca,Zhang:2014wra,Chang:2016ntp}. The SN1987A constraint places a limit on the mixing parameter.

@@

There are two processes to consider in order to estimate the rate of energy loss via $\Aprime$ bremsstrahlung. The first is  
proton-proton (pp) scattering with the bremsstrahlung of the dark photon off the proton; $p+p \rightarrow p+p+\Aprime$. 
The second is proton-neutron (pn) scattering with bremsstrahlung off of either the proton or the charged pion; 
$p+n \rightarrow p+n+\Aprime$. We estimate the rates for these processes using the one-pion exchange (OPE) 
approximation for nucleon interactions. In the pp case, there are eight tree-level diagrams, 
with the emission of the $\Aprime$ from each of the external legs. One of these diagrams is shown 
in Fig.~\ref{fig:ppdiagram}; the remaining seven diagrams come from placing the radiated $\Aprime$ 
on each of the other three protons and then, for each of these, interchanging the outgoing momenta. 
For the pn case, there are five diagrams, four of which are analogous to the pp diagram shown in 
Fig.~\ref{fig:ppdiagram}. The fifth diagram, shown in Fig.~\ref{fig:npdiagram}, corresponds to 
emission of the $\Aprime$ from the exchanged, charged pion. \akl{Cameron, can you make the $A' \to \mathrm{A'}$ in figures 1 and 2?}

@@

\begin{figure*}[th]
\includegraphics[width=14cm]{compiledbounds_edited.pdf}
\caption{
Constraints on dark photon models. Our primary result is the excluded region 
in the $\varepsilon$-$m_{\mathrm{A'}}$ plane shaded in blue. The red curve is our attempts to reproduce exactly Ref.~\cite{dent_etal12}, while the black curve is Ref.~\cite{dent_etal12} but with the correct luminosity constraint.  The purple curve is using the SRA fixing the cross section using nuclear scattering data, while the green curve uses the soft radiation approximation of the OPE result.  \arz{The labels in the figure legend appear to be incorrect now.} \akl{In the legend, the red and black lines need to be switched, and the "Full excluded region" needs to be switched with the "SRA using OPE model".}
}
\label{fig:constraint}
\end{figure*}

@@

Finally, we note that the interior temperature and density within the progenitor of SN1987A remain 
uncertain at least at the level of tens of percent \citep[e.g.][]{perego15,farmer16}. This is well worth 
noting because the constraints one derives using the techniques in this work or similar techniques 
are quite sensitive to the properties of the SN1987A core, particularly the core temperature. Though 
we do not show this result explicitly \akl{Seems to me we should...}, we have replicated our calculations for shifts in core temperature 
of $\pm 5\, \mathrm{MeV}$ and find that the boundaries of the excluded region shift by a factor of $\sim 2-3$ at low 
mass ($m_{\mathrm{A}'} \lesssim T$ where $T = 30\, \mathrm{MeV}$ in our fiducial calculation, following 
work or previous authors) and considerably more at high mass. 
We conclude that the differences among theoretical techniques for 
computing the excluded region (OPE, EFT, SRA, ...) lead to uncertainties that are of the same order or 
smaller than the uncertainty in the excluded region induced by our limited knowledge of the 
properties of the SN1987A progenitor interior. The excluded regions we and other authors delineate 
should be considered to be uncertain by a factor or $\sim 3$. Refining the theoretical approach to this 
problem is therefore unwarranted until a breakthrough in our understanding of the interior of the 
SN1987A progenitor is achieved.
